\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-04-24 forwards-and-backwards.tex
.

%%%%%%%%%%
\paragraph{Allen-Cahn equation:}
Now take the Allen-Cahn equation:
\begin{align} \label{eqn:allen_cahn}
    \partial_t\varphi = \partial_{xx}\varphi + \varphi(1-\varphi)(2\varphi-1+s),
\end{align}
for a given $s \in (0,2)$.
Once again we have taken $r=\gamma=1$, but now the reaction term is bistable.
This equation can be used to model the motion of so-called 
hybrid zones in population genetics; see, for example,
\cite{barton:1979, gooding:2018, etheridge/gooding/letter:2022}.
This equation has an explicit travelling wave solution with speed $s$
and shape
\[ w^A(x) = (1+e^{x})^{-1} , \]
i.e., $\phi_t(x)=w^A(x-st)$ solves~(\ref{eqn:allen_cahn}).
Substituting $w^A$ in place of $w^F$ in~(\ref{eqn:fkpp_generator}),
we find that the generator of an ancestral lineage is now, 
\begin{align*}
\Lgen_2 f
    &=
    \partial_{xx}f
    + 
    2 \frac{\partial_xw^A}{w^A} \partial_xf
    +
    s \partial_x f\\
    \qquad &=
    \partial_{xx}f
    -
    2 \frac{e^x}{1+e^x} \partial_xf 
    + 
    s \partial_xf,
\end{align*}
so lineages in the tip are pushed leftwards into the bulk of the 
wave at a rate $s-2e^x/(1+e^{x})$.
The strength of this drift decreases exponentially as $x\to -\infty$.
The density of the speed measure for this diffusion is
$$
    m(x) \propto e^{sx}(1+e^x)^{-2},
$$
which is integrable, and so determines the unique stationary distribution.
Thus the position of the ancestral lineage relative to the wavefront will converge to a stationary 
distribution which is maximised away from the extreme tip of the wave.
This is consistent with~\cite{etheridge/penington:2022}, who consider an analogous
stochastic population model, although the stronger result there (that the genealogy
of a sample from behind the wavefront is approximately a Kingman coalecsent) requires
the stronger condition $s<1$.

@@

@@

\begin{proof}[of Corollary~\ref{cor:stationary_dist}]
For the moment, we will write $r(x)$ for $r(x, \eta)$
and similarly for $\gamma(x)$.
First, note that in this case
$$
    \Lgen f = \sigma^2 r \gamma \left(
        \Delta f
        + \grad ( 2 \log(\gamma \varphi) - h ) \cdot \grad f
    \right) ,
$$
and that
$$
    \int_{\IR^d} e^{H(x)} f(x) (\Delta + \grad H(x) \cdot \grad) g(x) dx
    =
    - \int_{\IR^d} e^{H(x)} \left\{
        \grad f(x) \cdot \grad g(x)
    \right\} dx .
$$
Therefore, by choosing $H(x) = 2 \log(\gamma(x) \varphi(x)) - h(x)$
and
$$ \pi(x)
    = e^{H(x)} / (\sigma^2 r(x) \gamma(x))
    = \frac{\gamma(x) \varphi(x)^2 e^{-h(x)}}{\sigma^2 r(x)} ,
$$
we have that
$$
    \int_{\IR^d} \pi(x) f(x) \Lgen g(x) dx
    =
    - \int_{\IR^d} e^{H(x)} \grad f(x) \cdot \grad g(x) dx .
$$
Since this is symmetric in $f$ and $g$, the process $Y$ is reversible with respect to $\pi$;
the constant factor of $\sigma^2$ does not affect the result.
\end{proof}

>>>>

%%%%%%%%%%
\paragraph{Allen-Cahn equation:}
Now take the Allen-Cahn equation:
\begin{align} \label{eqn:allen_cahn}
    \partial_t\varphi = \partial_{xx}\varphi + \varphi(1-\varphi)(2\varphi-1+s),
\end{align}
for a given $s \in (0,2)$.
Once again we have taken $r=\gamma=1$, but now the reaction term is bistable.
This equation can be used to model the motion of so-called 
hybrid zones in population genetics; see, for example,
\cite{barton:1979, gooding:2018, etheridge/gooding/letter:2022}.
This equation has an explicit travelling wave solution with speed $s$
and shape
\[ w^A(x) = (1+e^{x})^{-1} , \]
i.e., $\phi_t(x)=w^A(x-st)$ solves~(\ref{eqn:allen_cahn}).
Substituting $w^A$ in place of $w^F$ in~(\ref{eqn:fkpp_generator}),
we find that the generator of an ancestral lineage is now, 
\begin{align*}
\Lgen_2 f
    &=
    \partial_{xx}f
    + 
    2 \frac{\partial_xw^A}{w^A} \partial_xf
    +
    s \partial_x f\\
    \qquad &=
    \partial_{xx}f
    -
    2 \frac{e^x}{1+e^x} \partial_xf 
    + 
    s \partial_xf,
\end{align*}
so lineages in the tip are pushed leftwards into the bulk of the 
wave at a rate $s-2e^x/(1+e^{x})$.
The strength of this drift decreases exponentially as $x\to -\infty$.
The density of the speed measure for this diffusion is
$$
    m(x) \propto e^{sx}(1+e^x)^{-2},
$$
which is integrable, and so determines the unique stationary distribution.
% (agreeing with Lemma~\ref{cor:stationary_dist}).
Thus the position of the ancestral lineage relative to the wavefront will converge to a stationary 
distribution which is maximised away from the extreme tip of the wave.
This is consistent with~\cite{etheridge/penington:2022}, who consider an analogous
stochastic population model, although the stronger result there (that the genealogy
of a sample from behind the wavefront is approximately a Kingman coalecsent) requires
the stronger condition $s<1$.

@@

It is interesting to compare the stationary distribution we have obtained here
to the expression that we'd get by setting $\meanq = -\wavespeed$
and using Lemma~\ref{cor:stationary_dist},
i.e., by giving each offspring a mean displacement that offsets the motion of the wave.
In the Fisher-KPP and Allen-Cahn cases above, we get the same expressions,
but this is only the case because $r\equiv\gamma\equiv1$ in both.
For the PME with $\meanq = 1$ we have $\DG^* = \Delta + \grad$ 
and so the equation solved by the population density is
$$
    \partial_t \varphi = \Delta \varphi + 2 \varphi \grad \varphi + \varphi (1 - \varphi) ,
$$
which has a traveling wave solution of the same shape but moving at half the speed,
$\varphi(x,t) = w^{P}(x-t/2)$,
and a stationary distribution of the lineage relative to the wavefront of
$\pi(\xi) \propto e^{3\xi/2} (1 - e^{\xi/2})^3$.
% \comment{Details to back that last statement up, which we probably want to comment out?}
% (We can see that $w^P(x-t/2)$ is a traveling wave solution by using that
% $\partial_x w^P = -(1 - w^P)/2$;
% and then the generator of a linage in the moving frame is
% $\Lgen_{P'} = \varphi \partial_x^2 + (4 \partial_x \varphi + 1/2 + \varphi) \partial_x$,
% which has speed measure proportional to
% $\exp(\int^x 3(1/2 dx + d \log \varphi)) / \varphi$.)
% \begin{align*}
% m_{P'}(\xi)
%     &\propto \frac{1}{2(1-e^{\xi/2})} \exp\left(
%         \frac{3}{2} \int^\xi \left\{
%             1 - \frac{e^{x/2}}{1 - e^{x/2}}
%         \right\}dx
%     \right) .
% \end{align*}

@@

\begin{proof}[of Corollary~\ref{cor:stationary_dist}]
For the moment, we will write $r(x)$ for $r(x, \eta)$
and similarly for $\gamma(x)$.
First, note that in this case
$$
    \Lgen f = \sigma^2 r \gamma \left(
        \Delta f
        + \grad ( 2 \log(\gamma \varphi) - h/\sigma^2 ) \cdot \grad f
    \right) ,
$$
and that
$$
    \int_{\IR^d} e^{H(x)} f(x) (\Delta + \grad H(x) \cdot \grad) g(x) dx
    =
    - \int_{\IR^d} e^{H(x)} \left\{
        \grad f(x) \cdot \grad g(x)
    \right\} dx .
$$
Therefore, by choosing $H(x) = 2 \log(\gamma(x) \varphi(x)) - h(x)/\sigma^2$
and
$$ \pi(x)
    = e^{H(x)} / (\sigma^2 r(x) \gamma(x))
    = \frac{\gamma(x) \varphi(x)^2 e^{-h(x)/\sigma^2}}{\sigma^2 r(x)} ,
$$
we have that
$$
    \int_{\IR^d} \pi(x) f(x) \Lgen g(x) dx
    =
    - \int_{\IR^d} e^{H(x)} \grad f(x) \cdot \grad g(x) dx .
$$
Since this is symmetric in $f$ and $g$, the process $Y$ is reversible with respect to $\pi$;
the constant factor of $\sigma^2$ does not affect the result.
\end{proof}
